---
title: "Assignment 10: Data Scraping"
author: "Lexi Nelson"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
#load needed packages
library(tidyverse)
library(here); 
here() #pointing me to the Assignments folder where I am currently working
library(rvest)


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
#indicate the NC Water site URL to be scraped
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
#scrape the water system name
water_system_name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water_system_name

#scrape the contact person
contact_person <- webpage %>% 
  html_nodes("div+ table tr:nth-child(4) td:nth-child(2)") %>% 
  html_text()
contact_person

#scrape the ownership
ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

#scrape maximum day use for each month
MGD_month <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
MGD_month
#this is an incomplete list because I was stumped at how to figure out how to
#get the selector to select everything. 

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
#convert scraped data into a dataframe
df_withdrawals <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                             "WaterSystemName" = rep(water_system_name),
                             "ContactPerson" = rep(contact_person),
                             "Ownership" = rep(ownership),
                             "MaximumDailyWithdrawl" = as.numeric(MGD_month))


#5 
#create line plot of maximum daily withdrawlas across months
df_withdrawals <-
  ggplot(df_withdrawals, aes(x = Month, y = MaximumDailyWithdrawl)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  #making sure months show on x axis since I got points off for that last hw
  labs(
    title = "Maximum Daily Withdrawals Across Months",
    x = "Month",
    y = "Maximum Daily Withdrawal (MGD)"
  )
show(df_withdrawals)
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
#construct a scraping function
scrape.it <- function(the_year, the_PWSID){
  #Get the proper url
  the_url <- 
    paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', the_PWSID, '&year=',the_year)
  print(the_url)

  #Fetch the website
  the_website <- read_html(the_url)

  #Scrape the data
  #<enter code from above here>
  the_water_system_name <- the_website %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
  the_contact_person <- the_website %>% html_nodes('div+ table tr:nth-child(4) td:nth-child(2)') %>% html_text()
  the_ownership <- the_website %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
  the_MGD_month <- the_website %>% html_nodes('th~ td+ td') %>% html_text()
  
  #Convert to dataframe
  #<enter code from above here, adding a column for the month>
  the_df <- data.frame(
    Month = c(1,5,9,2,6,10,3,7,11,4,8,12),
    Water_System_Name = the_water_system_name,
    Contact_Person = the_contact_person,
    Ownership = the_ownership,
    Maximum_Daily_Withdrawal = the_MGD_month

  #Return the dataframe
  return(the_df)
}

# test the above function 
df2024 <- scrape.it(2024, "03-32-010")
show(df2024)

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
#use the function to pull max daily withdrawals in 2020
df2020 <- scrape.it(2020, "03-32-010")
show(df2020)
#show the structure of the dataframe 
glimpse(df2020)

#code to plot max daily withdrawals
df_withdrawals_2020 <-
  ggplot(df2020, aes(x = Month, y = Maximum_Daily_Withdrawal, group=1)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) + #fix thius
  #making sure months show on x axis since I got points off for that last hw
  labs(
    title = "Maximum Daily Withdrawals Across Months",
    x = "Month",
    y = "Maximum Daily Withdrawal (MGD)"
  )
show(df_withdrawals_2020)


```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
#use function to extract data for Cary
dfC2020 <- scrape.it(2020, "03-92-020")
show(dfC2020)
#show the structure of the dataframe 
glimpse(dfC2020)
#change MGD column to numeric because it caused problems
dfC2020$Maximum_Daily_Withdrawal <-
  as.numeric(gsub(",", "", dfC2020$Maximum_Daily_Withdrawal))

#combine with Durham data collected above in a plot
#first use rbind to combine the two dataframes
df_combined = rbind(df2020, dfC2020)
df_withdrawals_2020_2 <-
  ggplot(df_combined, aes(x = Month, y = Maximum_Daily_Withdrawal, group=Water_System_Name, color = Water_System_Name)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) + #fix thius
  #making sure months show on x axis since I got points off for that last hw
  labs(
    title = "Maximum Daily Withdrawals Across Months",
    x = "Month",
    y = "Maximum Daily Withdrawal (MGD)",
    color = "Municipality")
show(df_withdrawals_2020_2)

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:
>
