---
title: "Assignment 8: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Lexi Nelson"
date: "Spring 2026"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**

>NOTE: For this assignment, you can skip tests for any of the assumptions required to run the models. 

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#set up session
#load in packages
library(tidyverse); library(here); library(agricolae)
library(lubridate); library(dplyr); library(ggplot2) 
#dplyr to use complete.cases command later
here() #check working directory
#returns "/home/guest/ENV872/EDE_Spring2026/Assignments"
#import raw data file
raw_data <-read.csv(
  here("Data/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"))
class(raw_data$sampledate)
#set date columns to date objects
raw_data$sampledate <- as.Date(raw_data$sampledate, format = "%m/%d/%y")
#check that conversion to date object was successful
class(raw_data$sampledate)


#2
#build a ggplot theme and set as default
mytheme <- theme_classic(base_size = 14) +
  theme(
    axis.text = element_text(color = "black"),
    axis.title = element_text(size = 12),
    axis.line = element_line(linewidth = 0.8),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    legend.key = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  )

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across
al lakes (slope of depth vs mean July lake temperature is zero).
Ha: Mean lake temperature recorded during July changes with depth across all
lakes (slope of depth vs mean July lake temperature is nonzero).


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
#wrangle the dataset
NTL_LTER_wrangled <- 
  raw_data %>% 
  mutate(Month = month(sampledate)) %>% #parse date column, add month column
  filter(Month == 7) %>% #keep only dates in July
  select("lakename", "year4", "daynum","depth", "temperature_C") %>% 
  #keep only certain columns
  filter(complete.cases(.))


#5
#create scatter plot of temp by depth
temp_by_depth_plot <- NTL_LTER_wrangled %>%
  ggplot(aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point() +
  geom_smooth(aes(group = 1), method = "lm", color = "black") + 
  #add only one smoothed line showing linear model as per instructions
  scale_y_continuous(limits = c(0,35)) + #limit temp values from 0 to 35
  labs(
    title = "Temperature by Depth of Lake",
    x = "Depth (Meters)",
    y = "Temperature (degrees C)",
    color = "Lake Name")
#add labels, and color code points by lake
print(temp_by_depth_plot)



```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The distribution of points suggest a linear relationship between the two variables. It appears that temperature decreases with increasing depth. This trend is strongest at a depth of between 0-10 meters.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
#perform a linear regression
regression <- 
  lm(NTL_LTER_wrangled$temperature_C ~ NTL_LTER_wrangled$depth)
print(regression) #displays the coefficients

#show the *strength* of the relationship
cor.test(NTL_LTER_wrangled$temperature_C, NTL_LTER_wrangled$depth)
#prints a p value of < 0.05, indicating that the null hypothesis is not supported
#cor is -0.859, strong negative relationship
summary(regression)


```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The cor value given by the Pearson's product-moment correlation is 
-0.859, therefore the r squared value is 0.738. Therefore, 73.8% of variation in 
temperature can be explained by changes in depth. This finding is based on 9726 
degrees of freedom (df=9726), and the p-value is 2.2e-16, indicating that the 
relationship between temperature and depth is statistically significant and we 
should reject the null hypothesis that mean lake temperature does not change 
with depth.
Furthermore, the slope is -1.946. This means that for every 1m change in depth,
the temperature is expected to decrease by 1.946 degrees Celsius.



---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
#run AIC:
#determine which set of explanatory variables is best suited to predict temp

# create model object with predictors
TPAIC <- lm(temperature_C ~ lakename + year4 + daynum + depth, 
            data = NTL_LTER_wrangled)

# Stepwise model selection using AIC
TPAIC_step <- step(TPAIC)

# View selected model
summary(TPAIC_step)

#The model retained year4, daynum, and depth, indicating that all of these 
#variables are suited to prediuct temperature.
#The p value for all three of these variables is less than 0.05, as shown in the
#4th column.


#10
#run a multiple regression on the recommended set of variables
multiple_regression <- lm(data = NTL_LTER_wrangled, 
                          temperature_C ~ year4 + daynum + depth)
summary(multiple_regression)
#summary outputs a multiple R-squared value of 0.7412


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests we 
use to predict temperature in our multiple regression is lakename, year4, daynum, 
and depth. The AIC output shows that removing any of these increases AIC. 
Please note that in my multiple regression, I only included year4, daynum, and 
depth since this is how I interpreted the instructions in Question 9. Since our 
multiple r-squared value is 0.7797, we can conclude that 77.97% of the variation 
in temperature is explained by the model consisting of these four variables. 
This is a minor improvement over the model using only depth, because the r 
squared for that model was 0.738.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#create two sets of models to see whether lakes have different temps in July

# Format ANOVA as aov
ANOVA_aov <- aov(data = NTL_LTER_wrangled, temperature_C ~ lakename)
summary(ANOVA_aov)

#results:  p value below 0.05 soreject null hypothesis i.e. difference between a
#pair of group means is statistically significant

# Format ANOVA as lm
ANOVA_lm <- lm(data = NTL_LTER_wrangled, temperature_C ~ lakename)
summary(ANOVA_lm)

#results:  p value below 0.05 so reject null hypothesis i.e. there are 
#significant differences in temp between lakes



```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, due to the p values being below 0.05 in both models, we can
conclude that there is a significant difference in mean temperature among the 
lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.height = 3, fig.width = 5}
#14.
#create a graph that depicts temperature by depth
temperature_by_depth_plot <- ggplot(NTL_LTER_wrangled, aes(x = depth, 
    y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +   # 50% transparent points
  geom_smooth(method = "lm", se = FALSE) +  # regression line for each lake
  scale_y_continuous(limits = c(0, 35)) +   # y-axis limits
  labs(
    title = "Temperature vs Depth by Lake",
    x = "Depth (m)",
    y = "Temperature (°C)",
    color = "Lake Name") +
  theme(legend.position = "right")
show(temperature_by_depth_plot)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
#run Tukey's HSD test
TukeyHSD(ANOVA_aov)
#the results indicate that many lakes have different means, since the p adj is 
#below 0.05 for many of the comparisons.


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: The lakes that have the same mean temperature, statistically speaking, 
as Peter Lake are Paul Lake and Ward Lake. Peter Lake-Paul Lake has a p-value of
0.224 while Peter Lake-Ward Lake has a p value of 0.782. No single lake has a 
mean temperature that is statistically distinct from all the other lakes.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: Looking at Peter Lake and Paul Lake, we could also use a two sample t 
test to see whether they have distinct mean temperatures. A two sample t test
uses the hypothesis that the mean of two samples is equivalent.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#wrangle July data to include only records for Crampton Lake and Ward Lake
NTL_LTER_wrangled_2 <- 
  NTL_LTER_wrangled %>% 
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

Ttest <- t.test(temperature_C ~ lakename, data = NTL_LTER_wrangled_2)
print(Ttest)
```

>Answer: The test states that the mean in Crampton lake is 15.35 and the mean in
War Lake is 14.46, and provides a 95 percent confidence interval for the 
difference in means. Most importantly, the p-value is 0.2649 > 0.05, meaning we
cannot reject the null hypothesis. The p-value fails to reject our null 
hypothesis that the true difference in means between Crampton lake and Ward lake
is equal to zero. This is consistent with our answer to question 16that no 
single lake has a mean temperature that is statistically distinct from all the 
other lakes.
